Revisiting the hydrodynamic modulation of short waves by longer waves

### 2024-09-06 paper/paper.tex
Gravity modulation section

\title{Revisiting the hydrodynamic wave modulation from linear wave theory}

@@

More recently, \citep{peureux2021unsteady} asked whether the steady solutions
of short-wave modulation are realistic by simulating the full wave conservation
equations
They found that the short waves grow unsteadily due to the propagation of
longer waves, suggesting that the steady-state solutions may only be valid for
a few long-wave periods before the short waves begin to break due to excessive
steepness.
However, the unsteadiness of their solutions occurs in a specific scenario of
a uniform short-wave field that suddenly appears in a uniform long-wave train,
and the numerical simulations stabilize if the short-wave field is initialized
from existing steady solutions.
Nevertheless, their results do put into question whether the short-wave
modulation predicted by theory is generally steady over many long-wave periods
or not.

@@

\begin{equation}
\label{eq:k_short_lhs}
\widetilde{k} = k (1 + \varepsilon_L \cos{\psi})
\end{equation}
Coming from the conservation of wave crests, as opposed to the velocity potential
superposition approach, the L-HS solution requires two approximations in their
solution:
First, evaluating the long-wave velocity at $z = 0$ rather than $z = \eta_L$; and
second, further truncating the Taylor expansion series beyond the first order term.
These approximations cause an underestimate of the short-wave modulation
magnitude (Fig. \ref{fig:analytical_solutions_ak0.1}a).
At the wave crest, Eq. \ref{eq:k_short_exact} yields the wavenumber modulation
that is 16.9\%, 38.5\%, 66.4\%, and 104\% larger than that of \citet{longuet1960changes}
for $\varepsilon_L = 0.1, 0.2, 0.3, 0.4$, respectively.

@@

Long waves induce orbital accelerations, so the short waves that ride on their
surface move in an accelerated reference frame and experience effective
gravitational acceleration that is lower at the crests and higher in the troughs
\citep{longuet1986eulerian,longuet1987propagation}.
Further, a smaller but non-negligible effect is that of the centripetal force
due to the curvature of the long wave
\citep{phillips1981dispersion,zhang1990evolution}.
These effects are completely described by projecting the gravitational
acceleration from the coordinate that is perpendicular to the mean water surface
($z=0$) to that which is perpendicular to the long-wave surface ($z=\eta$),
and likewise for the long-wave orbital accelerations:

@@

\begin{equation}
\label{eq:local_slope}
\alpha = \tan^{-1}{\dfrac{\partial \eta}{\partial x}}
\end{equation}
where $\alpha$ is the local slope of the long-wave surface.
For a linear wave, the effective gravity is:

@@

Curvilinear effects on effective gravity can be neglected by setting $\alpha = 0$.

The first-order term in $a_L k_L$ is in phase with the long-wave elevation,
so the apparent gravity of the short waves is increased at the long-wave crests
and decreased in the troughs.
The second-order term in $a_L k_L$ is positive-definite and phase-shifted by
$\pi/2$ relative to the crests.
The second-order term thus increases the apparent gravity on the front and rear
faces of the long wave (Fig. \ref{fig:gravity_modulation}).

We can also derive the short-wave gravitational acceleration on the surface of
a non-linear Stokes wave with the elevation (to the third order in $\varepsilon_L$):

@@

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{figures/fig_effective_gravities.png}
\caption{
  (a) Analytical solutions for the effective gravitational acceleration
  modulation by long waves, as function of long-wave phase, for $\varepsilon_L = 0.2$.
  Black is for linear wave evaluated at $z=0$,
  blue and green are for linear and Stokes waves, respectively, evaluated at $z=\eta$ and without curvature effects,
  and orange and green are for linear and Stokes waves, respectively, evaluated at $z=\eta$ and with curvature effects.
  (b) As in (a) but normalized by the modulation of the linear wave evaluated at $z=0$.
}
\label{fig:effective_gravities}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/fig_effective_gravities_mean_max_min.png}
\caption{
  As Fig. \ref{fig:effective_gravities} but showing the mean (solid),
  maximum (dashed), and minimum (dotted) short-wave gravity modulation as
  function of long-wave steepness $\varepsilon_L$.
}
\label{fig:effective_gravities_mean_max_min}
\end{figure}

@@

\backsection[Funding]{
  Milan Curcic has been partially supported by the National Science Foundation
  under grant number AGS-1745384.
}

>>>>

\title{Revisiting the hydrodynamic modulation of short waves by longer waves}

@@

More recently, \citep{peureux2021unsteady} asked whether the steady solutions
of short-wave modulation are realistic by simulating the full wave conservation
equations.
They found that the short waves grow unsteadily due to the propagation of
longer waves, suggesting that the steady-state solutions may only be valid for
a few long-wave periods before the short waves begin to break due to excessive
steepness.
However, the unsteadiness of their solutions occurs in a specific scenario of
a uniform short-wave field that suddenly appears in a uniform long-wave train,
and the numerical simulations stabilize if the short-wave field is initialized
from existing steady solutions.
Nevertheless, their results do put into question whether the short-wave
modulation predicted by theory is generally steady over many long-wave periods
or not.

@@

\begin{equation}
\label{eq:k_short_lhs}
\widetilde{k} = k (1 + \varepsilon_L \cos{\psi})
\end{equation}
Coming from the conservation of wave crests, as opposed to the velocity potential
superposition approach, the L-HS solution requires two approximations in their
solution:
First, evaluating the long-wave velocity at $z = 0$ rather than $z = \eta_L$; and
second, further truncating the Taylor expansion series beyond the first order term.
These approximations cause an underestimate of the short-wave modulation
magnitude (Fig. \ref{fig:analytical_solutions_ak0.1}a).
At the wave crest, Eq. \ref{eq:k_short_exact} yields the wavenumber modulation
that is 16.9\%, 38.5\%, 66.4\%, and 104\% larger than that of \citet{longuet1960changes}
for $\varepsilon_L = 0.1, 0.2, 0.3, 0.4$, respectively.
Although not immediately obvious from the expressions, the modulated wavenumber
in Eq. \ref{eq:k_short_lhs} is conservative across the long-wave phase, whereas
the higher-order solution in Eq. \ref{eq:k_short_exact} is not.
This is due to the fact that combining the orbital velocities evaluated at
$z=\eta$ and the linearized conservation of wave crests in Eq. \ref{eq:wavenumber}
results in excess short-wave wavenumber being created.
The solution of the full Eq. \ref{eq:wavenumber} of course must be conservative
across the long-wave phase, as we will see later from the numerical simulations.

@@

Long waves induce orbital accelerations, so the short waves that ride on their
surface move in an accelerated reference frame and experience effective
gravitational acceleration that is lower at the crests and higher in the troughs
\citep{longuet1986eulerian,longuet1987propagation}.
Further, centripetal force due to the curvature of the long wave induces small
but non-negligible effects on the short-wave effective gravity
\citep{phillips1981dispersion,zhang1990evolution}.
These effects are completely described by projecting the gravitational
acceleration from the coordinate that is perpendicular to the mean water surface
($z=0$) to that which is perpendicular to the long-wave surface ($z=\eta$),
and likewise for the long-wave orbital accelerations:

@@

\begin{equation}
\label{eq:local_slope}
\alpha = \tan^{-1}{\dfrac{\partial \eta}{\partial x}}
\end{equation}
where $\alpha$ is the local slope of the long-wave surface.
For a linear long wave, the effective gravity is:

@@

Curvilinear effects on effective gravity can be removed altogether by setting
$\alpha = 0$.
In that case, Eq. \ref{eq:gravity_modulation_general} simplifies to:

\begin{equation}
\label{eq:gravity_modulation_linear_no_curvature}
\widetilde{g} = g \left( 1 - \varepsilon_L \cos{\psi} e^{\varepsilon_L \cos{\psi}} \right)
\end{equation}
Further, if we evaluate the orbital accelerations at the mean water level $z=0$,
this further simplifies to:

\begin{equation}
\label{eq:gravity_modulation_linear_no_curvature_mean_level}
\widetilde{g} = g \left( 1 - \varepsilon_L \cos{\psi} \right)
\end{equation}
which is the form used by \citet{peureux2021unsteady}, for example.

The general form of the gravity modulation in Eq. \ref{eq:gravity_modulation_general}
is convenient because it allows us to easily attribute the effects of the different
processes on the gravity modulation.
It also makes it straightforward to evaluate the gravity modulation for a different
form of the long-wave elevation, by evaluating the orbital accelerations and the
local elevation slope at the appropriate level.
For a third-order Stokes wave, for example, the elevation is:

@@

Let us now examine the contributions of the different terms to the short-wave
effective gravity modulation.
Fig. \ref{fig:effective_gravities}a shows the effective gravity modulation of
short-wave as a function of the long-wave phase for $\varepsilon_L = 0.2$, for
a variety of the gravity modulation forms.
This value of $\varepsilon_L$ allows easier visualization of the modulation
differences, while still being relatively moderate.
To the first order in $\varepsilon_L$, the gravity modulation is out of phase
with the long-wave elevation, with the short waves experiencing a lower effective
gravity on the long-wave crests and a higher effective gravity in the troughs.
Evaluating the orbital accelerations at $z=\eta$ causes a gravity reduction
around the crests ($\approx6\%$) and the troughs ($\approx 3\%$), and an
increase on the front and rear faces ($\approx 4\%$) of the long wave, due to
the exponential change in the orbital accelerations away from the mean water level
(Fig. \ref{fig:effective_gravities}b).
The difference between the linear and Stokes long waves at this steepness is
negligible.
The long-wave elevation curvature has a second-order effect in $\varepsilon_L$.
Naturally, there is no curvature exactly at the crests and troughs, and its
effects are largest at the front and rear faces of the long wave, where the
covariance between the local slope and the horizontal accelerations are greatest.

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/fig_effective_gravities.png}
\caption{
  (a) Analytical solutions for the effective gravitational acceleration
  modulation by long waves, as function of long-wave phase, for $\varepsilon_L = 0.2$.
  Black is for linear wave evaluated at $z=0$,
  blue and green are for linear and Stokes waves, respectively, evaluated at $z=\eta$ and without curvature effects,
  and orange and green are for linear and Stokes waves, respectively, evaluated at $z=\eta$ and with curvature effects.
  (b) As in (a) but normalized by the modulation of the linear wave evaluated at $z=0$.
}
\label{fig:effective_gravities}
\end{figure}

It is instructive to also examine the mean, maximum, and minimum values of the
effective gravity as function of long-wave steepness $\varepsilon_L$
(Fig. \ref{fig:effective_gravities_mean_max_min}).
The long-wave averaged effective gravity of short waves remains equal to the
resting value of $g$ for all values of $\varepsilon_L$, except if the effects
of curvature are considered, in which case the average effective gravity is
reduced.
This reduction, in case of a long Stokes wave, is $\approx 3\%$ for
$\varepsilon_L = 0.2$ and $\approx 14\%$ for $\varepsilon_L = 0.4$.
As the effective gravity minima and maxima occur at crests and troughs,
respectively, the curvature effects only affect the phase-averaged values but
not the extremes.
The effective gravity at the crest of the steepest ($\varepsilon_L = 0.4$) long
Stokes wave is reduced by $\approx 64\%$.

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/fig_effective_gravities_mean_max_min.png}
\caption{
  As Fig. \ref{fig:effective_gravities} but showing the mean (solid),
  maximum (dashed), and minimum (dotted) short-wave gravity modulation as
  function of long-wave steepness $\varepsilon_L$.
  The minimum and maximum values for the cases with curvature effects
  ($\alpha=\frac{\partial \eta}{\partial x}$) are equivalent to those without
  curvature effects ($\alpha=0$) and obscure them in the plot.
}
\label{fig:effective_gravities_mean_max_min}
\end{figure}

@@

\backsection[Funding]{
  Milan Curcic was partially supported by the National Science Foundation
  under grant number AGS-1745384.
}
